The Categorical Structure of Profunctor Optics

### 2017-10-24 paper.tex
Many tiny updates

The composition map
\begin{align*}
\int^{M \in \M} \left(\C(S, M A) \times \C(M A', S')\right) \times \int^{M \in \M} \left( \C(R, M S) \times \C(M S', R')\right) \to \int^{M \in \M} \C(R, M A) \times \C(M A', R')
\end{align*}
of preoptics is induced by the following morphisms for every $M, N \in \M$:
\begin{align*}
&\C(S, M A) \times \C(M A', S') \times \C(R, N S) \times \C(N S', R')\\
\xrightarrow{\text{functoriality}} \quad& \C(NS, NM A) \times \C(NM A', NS') \times \C(R, N S) \times \C(N S', R')\\
\xrightarrow{\text{composition}} \quad& \C(R, N M A) \times \C(N M A', R') \\
\xrightarrow{\copr_{NM}} \quad&\int^{M \in \M} \C(R, M A) \times \C(M A', R')
\end{align*}
In other words, let $\langle l, r \rangle : (R, R') \hto (S, S')$ and $\langle l', r' \rangle : (S, S') \hto (A, A')$ with $M$ the complement for $\langle l, r \rangle$. The composite is then: $\langle l', r' \rangle \circ \langle l, r \rangle = \langle (M l')l, r(Mr') \rangle$.

@@

\begin{proposition}
For any $\C$ and $\M$, there is a functor $(-, -)_* : \C \times \C^\op \to \PreOptic_\M$, which is given on objects by $(S, S')_* = (S, S')$ and for a morphism $(f, g) : (S, S') \to (A, A')$ by $(f, g)_* = \langle \epsilon_A f, g \epsilon_{A'}^{-1} \rangle$.
\end{proposition}
\begin{proof}
To check functoriality, suppose we have $(f, g) : (S, S') \to (A, A')$ and $(f', g') : (A, A') \to (B, B')$ in $\C \times \C^\op$. Then:
\begin{align*}
(f', g')_* \circ (f, g)_* 
&= \langle \epsilon_B f', g' \epsilon_{B'}^{-1} \rangle \circ \langle \epsilon_A f, g \epsilon_{A'}^{-1} \rangle \\
&= \langle (I(\epsilon_B f'))\epsilon_A f, g \epsilon_{A'}^{-1} (I(g' \epsilon_{B'}^{-1}))\rangle && \text{(By definition of $\circ$)}\\
&= \langle (I\epsilon_B) (I f')\epsilon_A f, g \epsilon_{A'}^{-1} (I g')(I\epsilon_{B'}^{-1})\rangle && \text{(Functoriality of action of $I$)}\\
&= \langle (I\epsilon_B) \epsilon_B f' f, g g' \epsilon_{B'}^{-1} (I\epsilon_{B'}^{-1})\rangle && \text{(Naturality of $\epsilon$)}\\
&= \langle (\lambda_I B) \epsilon_B f' f, g g' \epsilon_{B'}^{-1} (\lambda_I^{-1}B')\rangle && \text{(Unitality of action)} \\
&= \langle \epsilon_B f' f, g g' \epsilon_{B'}^{-1} (\lambda_I^{-1}B') (\lambda_I B') \rangle && \text{(Coend relation)}  \\
&= \langle \epsilon_B f'f, g g' \epsilon_{B'}^{-1} \rangle \\
&= (f'f, gg')_*
\end{align*}
\end{proof}
\todo{Is this functor monoidal?}

@@

Equationally, suppose we have representatives $(l, r) : (S \to M \otimes A, M \otimes A' \to S')$ and $(l', r') : (T \to N \otimes B, N \otimes B' \to T')$. Then:
\begin{align*}
\langle l, r \rangle \otimes \langle l', r' \rangle &:= \langle (M \otimes s_{A,N} \otimes B)(l \otimes l'), (r \otimes r')(M \otimes s_{A',N} \otimes B') \rangle
\end{align*}

Now we check functoriality. Suppose we have:
\begin{align*}
\langle l_1, r_1 \rangle : (S_1, S_1') &\hto (S_2, S_2') \\
\langle l_2, r_2 \rangle : (S_2, S_2') &\hto (S_3, S_3') \\
\langle p_1, q_1 \rangle : (T_1, T_1') &\hto (T_2, T_2') \\
\langle p_2, q_2 \rangle : (T_2, T_2') &\hto (T_3, T_3')
\end{align*}
with complements $M_1, M_2, N_1, N_2$ respectively. Then: \todo {(This is a mess. String diagram?)}
\begin{align*}
&(\langle l_2, r_2 \rangle \otimes \langle p_2, q_2 \rangle) \circ (\langle l_1, r_1 \rangle \otimes \langle p_1, q_1 \rangle) \\
= \; &\langle (M_2 \otimes s_{S_3,N_2} \otimes T_3)(l_2 \otimes p_2), (r_2 \otimes q_2)(M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
&\circ \langle (M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), (r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2') \rangle \\
= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes N_1 \otimes l_2 \otimes p_2)(M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), \\
&\;(r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2')(M_1 \otimes N_1 \otimes r_2 \otimes q_2)(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)(M_1 \otimes  l_2 \otimes N_1 \otimes p_2)(l_1 \otimes p_1), \\
&\;(r_1 \otimes q_1)(M_1 \otimes r_2 \otimes N_1 \otimes q_2)(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1), \\
&\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
= \; & \text{\todo{(Move all structure isomorphisms to one side of the coend, then appeal to coherence)}} \\
= \; &\langle (M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S_3} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1) ,  \\
&\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S'_3} \otimes T'_3) \rangle \\
= \; &\langle (M_1 \otimes l_2)l_1, r_1(M_1 \otimes r_2) \rangle \otimes \langle (N_1 \otimes p_2)p_1, q_1(N_1 \otimes q_2) \rangle \\
= \; &(\langle l_2, r_2 \rangle  \circ \langle l_1, r_1 \rangle) \otimes (\langle p_2, q_2 \rangle \circ \langle p_1, q_1 \rangle)
\end{align*}

@@

\todo{Once we know that $(-, -)_*$ is monoidal, the pentagon and triangle axioms should lift straight from $\C$ and we're done}

\end{proof}

@@

\subsection{Optics}
For a fixed $M \in \M$, there is a map $\C(S, M A) \times \C(M A, S) \to \C(S, S)$ given by composition. This induces a map $outside : \PreOptic_\M((S, S), (A, A)) \to \C(S, S)$. By composition in the other direction we have a map $\C(S, M A) \times \C(M A, S) \to \C(M A, M A)$. Including this object $\C(M A, M A)$ into the coend $\int^{M \in \M} \C(M A, M A)$, we induce a map $inside : \PreOptic_\M((S, S), (A,A)) \to \int^{M \in \M} \C(M A, M A)$. 
\begin{definition}
An \emph{optic} from $S$ to $A$ is a preoptic $p$ from $(S, S)$ to $(A, A)$ such that $outside(p) = \id_S$, and $inside(p) = \langle t A \rangle$ for some $t : M \to M \in \M$.
\end{definition}

\begin{proposition}
\label{prop-onthenose}
If $p : (S,S) \hto (A,A)$ is an optic, then there exists a representative $p = \langle l, r \rangle$ such that $rl = \phi A$ on the nose for some $\phi : M \to M$.
\end{proposition}
\begin{proof}
\todo{This is painful but seems important. There may be a more general principle at work here but I couldn't see it.}

@@

Lenses are the ur-example of a class of optics.

@@

As remarked in the introduction \todo{(todo: remark in the introduction)}, the preoptics for this action are exactly pairs of $\fget$ and $\fput$ functions.
\begin{align*}
\PreOptic_\times((S, S), (A, A)) &= \int^{M \in \M} \C(S, M \times A) \times \C(M \times A, S) \\
&\cong \int^{M \in \M} \C(S, M) \times \C(S, A) \times \C(M \times A, S) && \text{(universal property of product)} \\
&\cong \C(S, A) \times \C(S \times A, S) && \text{(co-Yoneda)}
\end{align*}

@@

\begin{proposition}
Suppose there exists a map $x : 1 \to A$. Then if $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then the corresponding $p : S \hto A$ is a prelens.
\end{proposition}
\begin{proof}
Recall that the corresponding optic $p : S \hto A$ has representative $\langle [\id_s, \fget], \fput \rangle$.

Let $\fput_x : S \to S$ be the composite $S \to S \times 1 \xrightarrow{x} S \times A \xrightarrow{\fput} S$. The commutative diagram
\[
\begin{tikzcd}
S \times A \ar[r, "\fput_x \times A "] \ar[d, "lr", swap] & S \times A \ar[d, "\fput_x \times A"] \ar[dl, "lr"] \\
S \times A \ar[r, "\fput_x \times A", swap] & S \times A
\end{tikzcd}
\]
is a witness that $\langle lr \rangle = \langle \fput_x \times A \rangle$ in $\int^{M \in \M} \C(M A, M A)$.
\todo{Is there some property of $\times$ used here we can abstract?}
\end{proof}

@@

\begin{proposition}
Suppose $\C$ has finite limits and that there is a morphism $x : 1 \to A$. Then if $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then there exists $C \in \C$ with $S \cong C \times A$.
\end{proposition}
\begin{proof}
\todo{TODO: This isn't really what we want to show, we want to start from an optic, not from something obeying the $\fget/\fput$ laws.}

Set $C$ to be the pullback of $\fget$ along $x$, so there is a map $i : C \to S$ with $\fget \, i = x !_C$. There is also a map $j : S \to C$ induced by the following diagram:
\[
\begin{tikzcd}
S \ar[ddr, bend right = 20] \ar[dr, "j", dashed] \ar[r, "{[\id_S, x !_S]}"] & S \times A \ar[dr, "\fput"] & \\
& C \ar[r, "i"] \ar[d] \arrow[dr, phantom, "\lrcorner", very near start] & S \ar[d, "\fget"] \\
& 1\ar[r, "x", swap] & A
\end{tikzcd}
\]
which commutes by the $\fput\fget$ law. Note that $ji = \id_C$ by the universal property of pullbacks. The claim is that $\fput (i \times A) : C \times A \to S$ and $[j,\fget] : S \to C \times A$ are mutual inverses. This is easily checked:
\begin{align*}
\fput (i \times A)[j,\fget] &= \fput [ij,\fget] \\
&= \fput [\fput [\id_S, x!_S],\fget] && \text{(by definition of $j$)} \\
&= \fput [\id_S,\fget] && \text{(by $\fput\fput$)} \\
&= \id_S && \text{(by $\fget\fput$)}
\intertext{and}
[j,\fget]\fput (i \times A) &= [j\fput (i \times A),\fget\,\fput (i \times A)] && \text{(by universal property of product)} \\
&= [j\fput (i \times A), \pi_2 (i \times A)] && \text{(by $\fput\fget$)} \\
&= [j\fput (i \times A), \pi_2] && \\
&= [jij\fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S,x !_S] \fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S, x !_S] \pi_1 (i \times A), \pi_2] && \text{(by $\fput\fput$ \todo{todo: detail})}\\
&= [jij \pi_1 (i \times A), \pi_2] && \\
&= [jiji \pi_1, \pi_2] && \\
&= [\pi_1, \pi_2] && \\
&= \id_{C \times A}
\end{align*}

Additionally, the commutative diagram
\[
\begin{tikzcd}
C \times A \ar[r, "i \times A"] \ar[d, "\fput (i \times A)", swap] & S \times A \ar[d, "\fput"] \\
S \ar[r, equal] \ar[d, "{[j,\fget]}", swap] & S \ar[d, "{[\id_S, \fget]}"]  \\
C \times A \ar[r, "i \times A", swap] & S \times A
\end{tikzcd}
\]
is a witness that $\langle \fput (i \times A), [j,\fget] \rangle = \langle \fput, [\id_S, \fget] \rangle$ as elements of $\PreOptic_\M((S, S), (A, A))$.
\end{proof}

@@

Just as lenses are constructed by two concrete maps $\fget : S \to A$ and $\fput : S \times A \to S$, prisms are constructed by two maps $\freview : A \to S$ and $\fmatching : S \to S \sqcup A$. These names are taken from the Haskell \texttt{lens} library.
\begin{align*}
\PreOptic_\sqcup((S, S), (A, A)) &= \int^{M \in \M} \C(S, M \sqcup A) \times \C(M \sqcup A, S) \\
&\cong \int^{M \in \M} \C(S, M \sqcup A) \times \C(M, S) \times \C(A, S) && \text{(universal property of coproduct)} \\
&\cong \C(S, S \sqcup A) \times \C(A, S) && \text{(co-Yoneda)}
\end{align*}
So, if we are given a pre-prism $\langle l, r \rangle : (S, S) \hto (A, A)$ then associated $\freview$ and $\fmatching$ morphisms are given by $\freview = r \inr$ and $\fmatching = (r\inl \sqcup \id_A)l$

The laws for prisms are the obvious duals to the lens laws. 
\begin{align*}
\fmatching \; \freview &= \inr \\
[\id_S, \freview] \fmatching &= \id_S \\
(\fmatching \sqcup \id_A) \fmatching &= \mathrm{in}_{1,3} \, \fmatching
\end{align*}
\todo{
\begin{proposition}
When $\C = \Set$, and probably more generally, the third law is implied by the other two.
\end{proposition}
}

>>>>

The composition map
\begin{align*}
\int^{M \in \M} \left(\C(S, M A) \times \C(M A', S')\right) \times \int^{M \in \M} \left( \C(R, M S) \times \C(M S', R')\right) \to \int^{M \in \M} \C(R, M A) \times \C(M A', R')
\end{align*}
of preoptics is induced by the following morphisms for every $M, N \in \M$:
\begin{align*}
&\C(S, M A) \times \C(M A', S') \times \C(R, N S) \times \C(N S', R')\\
\to \,& \C(NS, NM A) \times \C(NM A', NS') \times \C(R, N S) \times \C(N S', R') && \text{(functoriality of action of $N$)} \\
\to \,& \C(R, N M A) \times \C(N M A', R') && \text{(composition)} \\
\to \,& \C(R, (N \otimes M) A) \times \C((N \otimes M) A', R') && \text{(composition with $\mu_{M,N,A}$)} \\
\to \,&\int^{M \in \M} \C(R, M A) \times \C(M A', R') && \text{($\copr_{N \otimes M}$)}
\end{align*}
In other words, let $\langle l, r \rangle : (R, R') \hto (S, S')$ and $\langle l', r' \rangle : (S, S') \hto (A, A')$ with $M$ the complement for $\langle l, r \rangle$. The composite is then: $\langle l', r' \rangle \circ \langle l, r \rangle = \langle \mu_{M,N,A}(M l')l, r(Mr')\mu^{-1}_{M,N,A'} \rangle$.

@@

\begin{proposition}
For any $\C$ and $\M$, there is a functor $(-, -)_* : \C \times \C^\op \to \PreOptic_\M$, which is given on objects by $(S, S')_* = (S, S')$ and for a morphism $(f, g) : (S, S') \to (A, A')$ by $(f, g)_* = \langle \epsilon_A f, g \epsilon_{A'}^{-1} \rangle$.
\end{proposition}
\begin{proof}
This preserves identities, as the identity on an object $(S, S')$ in $\PreOptic_\M$ is defined to be exactly $\langle \epsilon_S, \epsilon_{S'}^{-1} \rangle$.

To check functoriality, suppose we have $(f, g) : (S, S') \to (A, A')$ and $(f', g') : (A, A') \to (B, B')$ in $\C \times \C^\op$. Then:
\begin{align*}
(f', g')_* \circ (f, g)_* 
&= \langle \epsilon_B f', g' \epsilon_{B'}^{-1} \rangle \circ \langle \epsilon_A f, g \epsilon_{A'}^{-1} \rangle \\
&= \langle (I(\epsilon_B f'))\epsilon_A f, g \epsilon_{A'}^{-1} (I(g' \epsilon_{B'}^{-1}))\rangle && \text{(By definition of $\circ$)}\\
&= \langle (I\epsilon_B) (I f')\epsilon_A f, g \epsilon_{A'}^{-1} (I g')(I\epsilon_{B'}^{-1})\rangle && \text{(Functoriality of action of $I$)}\\
&= \langle (I\epsilon_B) \epsilon_B f' f, g g' \epsilon_{B'}^{-1} (I\epsilon_{B'}^{-1})\rangle && \text{(Naturality of $\epsilon$)}\\
&= \langle (\lambda_I B) \epsilon_B f' f, g g' \epsilon_{B'}^{-1} (\lambda_I^{-1}B')\rangle && \text{(Unitality of action)} \\
&= \langle \epsilon_B f' f, g g' \epsilon_{B'}^{-1} (\lambda_I^{-1}B') (\lambda_I B') \rangle && \text{(Coend relation)}  \\
&= \langle \epsilon_B f'f, g g' \epsilon_{B'}^{-1} \rangle \\
&= (f'f, gg')_*
\end{align*}
\end{proof}
\todo{Is this functor monoidal?}

@@

Equationally, suppose we have representatives $(l, r) : (S \to M \otimes A, M \otimes A' \to S')$ and $(l', r') : (T \to N \otimes B, N \otimes B' \to T')$. Then:
\begin{align*}
\langle l, r \rangle \otimes \langle l', r' \rangle &:= \langle (M \otimes s_{A,N} \otimes B)(l \otimes l'), (r \otimes r')(M \otimes s_{A',N} \otimes B') \rangle
\end{align*}
For future convenience, note that 
\begin{align*}
\langle l, r \rangle \otimes (T, T') &= \langle (M \otimes s_{A,I} \otimes T)(l \otimes \lambda_T), (r \otimes \lambda^{-1}_{T'})(M \otimes s_{A',I}\otimes T') \rangle \\
&= \langle (\lambda^{-1}_M \otimes A \otimes T)(M \otimes s_{A,I} \otimes T)(l \otimes \lambda_T), (r \otimes \lambda^{-1}_{T'})(M \otimes s_{A',I}\otimes T')(\lambda_M \otimes A' \otimes T') \rangle \\
&= \langle l \otimes T, r \otimes T' \rangle
\end{align*}
Now we check functoriality. Suppose we have:
\begin{align*}
\langle l_1, r_1 \rangle : (S_1, S_1') &\hto (S_2, S_2') \\
\langle l_2, r_2 \rangle : (S_2, S_2') &\hto (S_3, S_3') \\
\langle p_1, q_1 \rangle : (T_1, T_1') &\hto (T_2, T_2') \\
\langle p_2, q_2 \rangle : (T_2, T_2') &\hto (T_3, T_3')
\end{align*}
with complements $M_1, M_2, N_1, N_2$ respectively. Then: \todo {(This is a mess. String diagram?)}
\begin{align*}
&(\langle l_2, r_2 \rangle \otimes \langle p_2, q_2 \rangle) \circ (\langle l_1, r_1 \rangle \otimes \langle p_1, q_1 \rangle) \\
= \; &\langle (M_2 \otimes s_{S_3,N_2} \otimes T_3)(l_2 \otimes p_2), (r_2 \otimes q_2)(M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
&\circ \langle (M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), (r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2') \rangle \\
= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes N_1 \otimes l_2 \otimes p_2)(M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), \\
&\;(r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2')(M_1 \otimes N_1 \otimes r_2 \otimes q_2)(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)(M_1 \otimes  l_2 \otimes N_1 \otimes p_2)(l_1 \otimes p_1), \\
&\;(r_1 \otimes q_1)(M_1 \otimes r_2 \otimes N_1 \otimes q_2)(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
= \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1), \\
&\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
= \; & \text{\todo{(Add a $s_{M_2, N_1}$ to either side of the coend, then appeal to coherence)}} \\
= \; &\langle (M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S_3} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1) ,  \\
&\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S'_3} \otimes T'_3) \rangle \\
= \; &\langle (M_1 \otimes l_2)l_1, r_1(M_1 \otimes r_2) \rangle \otimes \langle (N_1 \otimes p_2)p_1, q_1(N_1 \otimes q_2) \rangle \\
= \; &(\langle l_2, r_2 \rangle  \circ \langle l_1, r_1 \rangle) \otimes (\langle p_2, q_2 \rangle \circ \langle p_1, q_1 \rangle)
\end{align*}

@@

Note that because $(S, S')_* = (S, S')$, required equations for $(-, -)_*$ to be a monoidal functor hold by definition (although we don't yet know that $\PreOptic_\otimes$ is monoidal). The pentagon and triangle axioms then hold in $\PreOptic_\otimes$, as they are the image of the same axioms in $\C \times \C^\op$ under $(-, -)_*$. The only remaining piece to verify is that these structure maps are natural in $\PreOptic_\otimes$.

Let us check that $\alpha_{(R, R'), (S, S'), (T, T')}$ is natural in $(R, R')$. Suppose $f : (Q, Q') \hto (R, R')$ has representative $(l, r) : (Q \to M \otimes R, M \otimes R' \to Q')$, then:
\begin{align*}
&\alpha_{(R, R'), (S, S'), (T, T')} \circ (f \otimes (S, S')) \otimes (T, T') \\
&= (\alpha_{R,S,T}, \alpha_{R',S',T'}^{-1})_*  \circ (\langle l, r \rangle \otimes (S, S')) \otimes (T, T') \\
&= \langle\lambda_{R \otimes (S \otimes T)}\alpha_{R,S,T}, \alpha_{R',S',T'}^{-1} \lambda^{-1}_{(R' \otimes S') \otimes T'} \rangle  \circ \left\langle (l \otimes S) \otimes T, (r \otimes S') \otimes T' \right\rangle \\
&=\left\langle (M \otimes \lambda_{R \otimes (S \otimes T)}\alpha_{R,S,T})((l \otimes S) \otimes T), ((r \otimes S') \otimes T') (M \otimes \alpha_{R',S',T'}^{-1} \lambda^{-1}_{(R' \otimes S') \otimes T'} ) \right\rangle \\
&=\left\langle (M \otimes \alpha_{R,S,T})((l \otimes S) \otimes T), ((r \otimes S') \otimes T') (M \otimes \alpha_{R',S',T'}^{-1} )\right\rangle \\
&=\left\langle (l \otimes (S \otimes T))\alpha_{Q,S,T}, \alpha_{Q',S',T'}^{-1}(r \otimes (S' \otimes T')) \right\rangle \\
&= \langle (I \otimes l \otimes (S \otimes T)) \lambda_{Q \otimes (S \otimes T)}\alpha_{Q,S,T}, (I \times r \otimes (S' \otimes T')) \alpha_{Q',S',T'}^{-1} \lambda^{-1}_{Q' \otimes (S' \otimes T')}\rangle \\
&= \langle l \otimes (S \otimes T), r \otimes (S' \otimes T')\rangle \circ \left\langle \lambda_{Q \otimes (S \otimes T)}\alpha_{Q,S,T}, \alpha_{Q',S',T'}^{-1} \lambda^{-1}_{Q' \otimes (S' \otimes T')} \right\rangle \\
&= \langle l, r \rangle \otimes ((S, S') \otimes (T, T')) \circ (\alpha_{Q,S,T}, \alpha_{Q',S',T'}^{-1})_* \\
&= f \otimes ((S, S') \otimes (T, T')) \circ \alpha_{(Q, Q'), (S, S'), (T, T')} 
\end{align*}
\todo{There is some cheating going on here, composition actually adds an extra $\alpha$, but that wouldn't get in the way.}
\end{proof}

@@

There is some further useful structure in $\PreOptic_\otimes$. The set of costates $(S, S') \hto (I, I)$ is isomorphic to $\C(S, S')$:
\begin{align*}
\PreOptic_\otimes((S, S'), (I, I)) 
&= \int^{M \in \C} \C(S, M \otimes I) \times \C(M \otimes I, S') \\
&\cong \int^{M \in \C} \C(S, M) \times \C(M, S') \\
&\cong \C(S, S')
\end{align*}
The identity morphism $\C(S, S)$ gives us a preoptic $\varepsilon : (S, S) \to (I, I)$ that we call the \emph{counit}.

States are not as easy to describe in general.

\begin{proposition}
Suppose the monoidal unit $I$ of $\C$ is terminal. Then the set of states $(I, I) \hto (A, A')$ is isomorphic to $\C(I, A)$.
\end{proposition}
\begin{proof}
\begin{align*}
\PreOptic_\otimes((I,I), (A,A')) 
&= \int^{M \in \C} \C(I, M \otimes A) \times \C(M \otimes A', I) \\
&\cong \int^{M \in \C} \C(I, M \otimes A)
\end{align*}
Because the interior of this coend is mute in the contravariant position, the coend is equal to the colimit of the functor $\C(I, - \otimes A) : \C \to \Set$. But $\C$ has a terminal object, so $\int^{M \in \C} \C(I, M \otimes A) \cong \C(I, I \otimes A) \cong \C(I, A)$.
\end{proof}

\subsection{Optics}
In this section we will focus only on preoptics of the form $p : (S,S) \hto (A, A)$. In what follows, we abbreviate $\PreOptic_\M((S, S), (A, A))$ as $\PreOptic_\M(S, A)$ and $p : (S, S) \hto (A, A)$ as $p : S \hto A$.

For a fixed $M \in \M$, there is a map $\C(S, M A) \times \C(M A, S) \to \C(S, S)$ given by composition. This induces a map $outside : \PreOptic_\M(S, A) \to \C(S, S)$. By composition in the other direction we have a map $\C(S, M A) \times \C(M A, S) \to \C(M A, M A)$. Including this object $\C(M A, M A)$ into the coend $\int^{M \in \M} \C(M A, M A)$, we induce a map $inside : \PreOptic_\M(S, A) \to \int^{M \in \M} \C(M A, M A)$. 
\begin{definition}
An preoptic $p : S \hto A$ is an \emph{optic} if $outside(p) = \id_S$, and $inside(p) = \langle t A \rangle$ for some $t : M \to M$ in $\M$.
\end{definition}

\begin{proposition}
\label{prop-onthenose}
If $p : S \to A$ is an optic, then there exists a representative $p = \langle l, r \rangle$ such that $rl = \phi A$ on the nose for some $\phi : M \to M$.
\end{proposition}
\begin{proof}
\todo{This is painful but seems important. There may be a more general principle at work here but I couldn't see it.}

@@

Lenses form the ur-example of a category of optics.

@@

As remarked in the introduction \todo{(todo: remark in the introduction)}, the preoptics for this action are exactly pairs of $\fget$ and $\fput$ functions.
\begin{align*}
\PreOptic_\times(S, A) &= \int^{M \in \M} \C(S, M \times A) \times \C(M \times A, S) \\
&\cong \int^{M \in \M} \C(S, M) \times \C(S, A) \times \C(M \times A, S) && \text{(universal property of product)} \\
&\cong \C(S, A) \times \C(S \times A, S) && \text{(co-Yoneda)}
\end{align*}

@@

\begin{proposition}
Suppose there exists a map $x : 1 \to A$. Then if $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then the corresponding $p : S \hto A$ is a lens.
\end{proposition}
\begin{proof}
Recall that the corresponding optic $p : S \hto A$ has as a representative $\langle [\id_s, \fget], \fput \rangle$.

Let $\fput_x : S \to S$ be the composite $S \to S \times 1 \xrightarrow{x} S \times A \xrightarrow{\fput} S$. The commutative diagram
\[
\begin{tikzcd}
S \times A \ar[r, "\fput_x \times A "] \ar[d, "lr", swap] & S \times A \ar[d, "\fput_x \times A"] \ar[dl, "lr"] \\
S \times A \ar[r, "\fput_x \times A", swap] & S \times A
\end{tikzcd}
\]
is a witness that $\langle lr \rangle = \langle \fput_x \times A \rangle$ in $\int^{M \in \M} \C(M A, M A)$.
\todo{Is there some property of $\times$ used here we can abstract? Having a map $1 \to A$ guarantees that $- \times A$ is faithful, is this the trick?}
\end{proof}

@@

\begin{proposition}
Suppose $\C$ has finite limits and that there is a morphism $x : 1 \to A$. Then if $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then there exists $C \in \C$ with $S \cong C \times A$.
\end{proposition}
\begin{proof}
Set $C$ to be the pullback of $\fget$ along $x$, so there is a map $i : C \to S$ with $\fget \, i = x !_C$. There is also a map $j : S \to C$ induced by the following diagram:
\[
\begin{tikzcd}
S \ar[ddr, bend right = 20] \ar[dr, "j", dashed] \ar[r, "{[\id_S, x !_S]}"] & S \times A \ar[dr, "\fput"] & \\
& C \ar[r, "i"] \ar[d] \arrow[dr, phantom, "\lrcorner", very near start] & S \ar[d, "\fget"] \\
& 1\ar[r, "x", swap] & A
\end{tikzcd}
\]
which commutes by the $\fput\fget$ law. Note that $ji = \id_C$ by the universal property of pullbacks. The claim is that $\fput (i \times A) : C \times A \to S$ and $[j,\fget] : S \to C \times A$ are mutual inverses. This is easily checked:
\begin{align*}
\fput (i \times A)[j,\fget] &= \fput [ij,\fget] \\
&= \fput [\fput [\id_S, x!_S],\fget] && \text{(by definition of $j$)} \\
&= \fput [\id_S,\fget] && \text{(by $\fput\fput$)} \\
&= \id_S && \text{(by $\fget\fput$)}
\intertext{and}
[j,\fget]\fput (i \times A) &= [j\fput (i \times A),\fget\,\fput (i \times A)] && \text{(by universal property of product)} \\
&= [j\fput (i \times A), \pi_2 (i \times A)] && \text{(by $\fput\fget$)} \\
&= [j\fput (i \times A), \pi_2] && \\
&= [jij\fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S,x !_S] \fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S, x !_S] \pi_1 (i \times A), \pi_2] && \text{(by $\fput\fput$ \todo{todo: detail})}\\
&= [jij \pi_1 (i \times A), \pi_2] && \\
&= [jiji \pi_1, \pi_2] && \\
&= [\pi_1, \pi_2] && \\
&= \id_{C \times A}
\end{align*}

Additionally, the commutative diagram
\[
\begin{tikzcd}
C \times A \ar[r, "i \times A"] \ar[d, "\fput (i \times A)", swap] & S \times A \ar[d, "\fput"] \\
S \ar[r, equal] \ar[d, "{[j,\fget]}", swap] & S \ar[d, "{[\id_S, \fget]}"]  \\
C \times A \ar[r, "i \times A", swap] & S \times A
\end{tikzcd}
\]
is a witness that $\langle \fput (i \times A), [j,\fget] \rangle = \langle \fput, [\id_S, \fget] \rangle$ as elements of $\PreOptic_\M(S, A)$.
\end{proof}

@@

Just as lenses are constructed by two concrete maps $\fget : S \to A$ and $\fput : S \times A \to S$, prisms are constructed by two maps $\freview : A \to S$ and $\fmatching : S \to S \sqcup A$. These names are taken from the Haskell \texttt{lens} library.
\begin{align*}
\PreOptic_\sqcup(S, A) &= \int^{M \in \M} \C(S, M \sqcup A) \times \C(M \sqcup A, S) \\
&\cong \int^{M \in \M} \C(S, M \sqcup A) \times \C(M, S) \times \C(A, S) && \text{(universal property of coproduct)} \\
&\cong \C(S, S \sqcup A) \times \C(A, S) && \text{(co-Yoneda)}
\end{align*}
So, if we are given a pre-prism $\langle l, r \rangle : (S, S) \hto (A, A)$ then associated $\freview$ and $\fmatching$ morphisms are given by $\freview = r \inr$ and $\fmatching = (r\inl \sqcup \id_A)l$

The laws for prisms are the obvious duals to the lens laws. 
\begin{align*}
\fmatching \; \freview &= \inr \\
[\id_S, \freview] \fmatching &= \id_S \\
(\fmatching \sqcup \id_A) \fmatching &= \mathrm{in}_{1,3} \, \fmatching
\end{align*}

\todo{
\begin{proposition}
When $\C = \Set$, and probably more generally, the third law is implied by the other two.
\end{proposition}
}
